Iterative construction of replicated designs based on Sobol' sequences

### 2016-06-08 crass/crassnew.tex
crass + article changes

\selectlanguage{francais}
% Text of abstract in French
\noindent{\bf R\'esum\'e} \vskip 0.5\baselineskip \noindent
{ \bf Construction de plans r\'epliqu\'es \`a partir de s\'equences de Sobol'.}
%{\bf Plans r\'epliqu\'es \`a base de s\'equences de Sobol' pour estimer les effets principaux des param\`etres d'un mod\`ele. }
Dans l'objectif d'estimer les effets principaux des param\`etres d'un mod\`ele, nous proposons d'\'etudier deux approches pour construire it\'erativement des plans r\'epliqu\'es \`a partir de s\'equences de Sobol'. Les propri\'et\'es de remplissement de l'espace des plans construits sont \'etudi\'ees sur la base de deux crit\`eres.
{\it Pour citer cet article~: L. Gilquin, Ll. A. Jim\'enez Rugama, E. Arnaud, F. J. Hickernell, H. Monod, C. Prieur, C. R. Acad. Sci.
Paris, Ser. I 340 (2005).}

\end{abstract}
\end{frontmatter}

\selectlanguage{english}

\section{Introduction}
Mathematical models often involve a substantial number of poorly known parameters. The effect of these parameters on the output of the model can be assessed through sensitivity analysis. Global sensitivity analysis methods are useful tools to identify the parameters having the most influence on the output. A well known approach is the variance based method introduced by Sobol' in \cite{Sobol2}. This method estimates sensitivity indices called Sobol' indices that summarize the influence of each model input. Among all Sobol' indices, one can distinguish first-order indices that estimate the main effect of each input.

@@

\subsection{Preliminaries}

Digital nets and sequences were first introduced by Niederreiter \cite{Niederreiter} in the numerical integration framework to define good uniformly distributed points in $\cube$. They can also appear in the literature as digital $(t,m,s)$-nets and digital $(t,s)$-sequences, or simply $(t,m,s)$-nets and $(t,s)$-sequences. Sobol' and Niederreiter-Xing sequences are two examples of digital sequences detailed in \cite{Sobol1} and \cite{NiederreiterXing}.

@@

Consider the generating matrices $C_1,\dots,C_s$, and $C_1^m,\dots,C_s^m$ their upper left corner blocks of size $m\times m$. Although $C_1,\dots,C_s$ are of infinite size, one only requires the knowledge of $C_1^m,\dots,C_s^m$ to construct the first $2^m$ points: for each $i=0,\dots,2^m-1$, the point $\vx_i = (x_{i,1},\dots,x_{i,s})^\intercal$ of the sequence is obtained dimension-wise by:
\begin{equation}
\label{dig.net.eq.}
(x_{i,j,1},\dots,x_{i,j,m})^\intercal = C_j^{m} \vi,\qquad j= 1,\dots,s\, ,
\end{equation}
where $x_{i,j} = \sum_{k \geq 1}^m x_{i,j,k}2^{-k}$ is the binary expansion of $x_{i,j}$ and $\vi = (i_{0},\dots,i_{m-1})^\intercal$ is the vector obtained from the binary expansion of $i= \sum_{k \geq 0}^{m-1} i_{k}2^{k}$. All matrix operations are performed in $\mathbb{F}_2$. Below we provide an example on how to compute $x_{11,1}=0.8125$ and $x_{9,2}=0.4375$,
\begin{equation*}
\begin{pmatrix}
x_{11,1,1} \\
x_{11,1,2} \\
x_{11,1,3} \\
x_{11,1,4}
\end{pmatrix} = 
\overbrace{
\begin{pmatrix}
1 & 0 & 0 & 0 \\
0 & 1 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1
\end{pmatrix}}^{C_1^4}
\overbrace{
\begin{pmatrix}
1 \\
1 \\
0 \\
1
\end{pmatrix}}^{\bvec{11}}=
\begin{pmatrix}
1 \\
1 \\
0 \\
1
\end{pmatrix}
,\qquad
\begin{pmatrix}
x_{9,2,1} \\
x_{9,2,2} \\
x_{9,2,3} \\
x_{9,2,4}
\end{pmatrix} = 
\overbrace{\begin{pmatrix}
1 & 1 & 1 & 1 \\
0 & 1 & 0 & 1 \\
0 & 0 & 1 & 1 \\
0 & 0 & 0 & 1
\end{pmatrix}}^{C_2^4}
\overbrace{\begin{pmatrix}
1 \\
0 \\
0 \\
1
\end{pmatrix}}^{\bvec{9}} =
\begin{pmatrix}
0 \\
1 \\
1 \\
1
\end{pmatrix}
\end{equation*}
\bigskip

@@

\begin{lemma}\label{Sobol_replicated}
All Sobol' sequences are digitally replicated of order 1.
\end{lemma}
\begin{pf*}{Proof.}
Consider any two $s$-dimensional Sobol' sequences generated by $C_1,\dots,C_s$ and $C'_1,\dots,C'_s$ respectively. Since any two matrices $C^{m}_j$ and ${C'}^{m}_j$ are square and full rank, the operations $\vi \mapsto C^{m}_j\vi$ and $\vi \mapsto {C'}^{m}_j\vi$ are one-to one and onto for all $i=0,\dots,2^m-1$. Therefore, they generate the same point sets.
\end{pf*}

@@

Two replicated point sets of order 1 $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ can be constructed using two $s$-dimensional Sobol' sequences. We note $C_1,\dots,C_s$ the generating matrices used to generate $\mathcal{P}_\ell$ and ${C'}_1,\dots,{C'}_s$ those used to generate $\mathcal{P}'_\ell$. All these matrices need to be different.
% Our choice of the $2s$ generating matrices $C_1,\dots,C_s,{C'}_1,\dots,{C'}_s$ follows the results obtained by Joe and Kuo in \cite{Kuo}. These two authors proposed a set of generating matrices optimized based on the $t$-value of pairwise projections of the corresponding digital sequence. Here, we select $2s$ generating matrices from this set to generate our two designs $\mathcal{P}$ and $\mathcal{P}'$.
\bigskip

@@

With the multiplicative approach, the size of designs $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ is multiplied by $2$ each time $\ell$ is increased by one. This growth rate may be inadequate for some applications. The additive approach presented in this section is attractive due to a slower size growth. Given an initial choice of $r\geq 1$ that specifies the size of $B_0$ and ${B'}_0$, only $|B_0|=2^r$ points are added to both designs at each step.

The main drawback of this approach is that $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ do not possess the structure of a Sobol' sequence anymore when $m\geq 1$. Still, both designs display good space-filling properties as shown in the next section.

@@

The two replicated point sets $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ constructed with the additive approach are iteratively refined with $B_\ell$ and ${B'}_\ell$. Each $B_\ell$ and ${B'}_\ell$ are $(t,r,s)$-nets.

First, $B_0=\{\vx_i^{(0)}\}_{i=0}^{2^r-1}$ and ${B'}_0=\{{\vx'}_i^{(0)}\}_{i=0}^{2^r-1}$ are set to be the first $2^r$ Sobol' points of two $s$-dimensional sequences. Then, $B_\ell$ (resp. ${B'}_\ell$), for $\ell \geq 1$, is obtained from $B_0$ (resp. ${B'}_0$) by carrying out digital shifts and scrambling operations. These operations guarantee that both $B_\ell$ and ${B'}_\ell$ inherit the $(t,r,s)$-net structure of the initial sets $B_0$ and ${B'}_0$. We detail below the construction of $B_\ell$ and ${B'}_\ell$.
\bigskip

@@

The left multiplications carried out in Equations (\ref{dig.shift}) and (\ref{ibinom.scrambling}) by matrices $L$ and $L'$ are referred as random linear scrambling \cite{Matousek} while the additions by vectors $\ve_j$ and ${\ve'}_j$ are referred as digital shifts. There are exactly $2^r$ different choices for vectors $\ve_j$ and ${\ve'}_j$ and $2^{r(r-1)/2}$ choices for matrices $L$ and $L'$.

The vectors $\ve_j$ (resp. ${\ve}'_j$) are selected to guarantee that the set $\{\ve_1,\dots,\ve_s\}$ (resp. $\{{\ve}'_1,\dots,{\ve}'_s\}$) is different at each step $\ell$. There exists exactly $2^{r s}$ combinations of this set which leaves us with a large panel of choice.

@@

\begin{lemma}
Using the additive construction, $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ are two replicated designs of order $1$ for $\ell \geq 0$.
\end{lemma}

\begin{pf*}{Proof.}
For each $\ell\geq 1$, digital shift and scrambling operations carried out in equations (\ref{dig.shift}) and (\ref{ibinom.scrambling}) are bijections from $\mathbb{F}^r_2$ to $\mathbb{F}^r_2$. Hence, same coordinates of $B_\ell$ and ${B'}_\ell$ will contain the same set of $2^r$ values.
\end{pf*}

@@

We denote by design $A$ (resp. design $B$) the design constructed with the multiplicative (resp. additive) approach. Values of maximin and discrepancy criteria are computed for designs $A$ and $B$.

For design $A$, the design constructed is refined in four steps, starting from $2^8$ points. For design $B$, we chose a starting value $r=8$ and refine the design $15$ times. To highlight the performance of design $B$, the results are compared to those obtained with a Latin Hypercube design. Average results over $100$ repeats are presented in Figure \ref{results.crit}. The x-axis represent the growing size of the design from $2^8$ points to $2^{12}$ points.

As expected design $A$ performs the best. However, we observe that design $B$ outperforms a classic Latin Hypercube design in both criteria while allowing $15$ construction steps instead of $4$ for design $A$.

\begin{figure}
\label{results.crit}
\includegraphics[scale=0.43]{discrep.pdf}
\hspace*{0.1cm}
\includegraphics[scale=0.43]{mindist.pdf}
\end{figure}

>>>>

\selectlanguage{francais}
% Text of abstract in French
\noindent{\bf R\'esum\'e} \vskip 0.5\baselineskip \noindent
{ \bf Construction de plans r\'epliqu\'es \`a partir de s\'equences de Sobol'.}
%{\bf Plans r\'epliqu\'es \`a base de s\'equences de Sobol' pour estimer les effets principaux des param\`etres d'un mod\`ele. }
Dans l'objectif d'estimer les effets principaux des param\`etres d'un mod\`ele, nous proposons d'\'etudier deux approches pour construire it\'erativement des plans r\'epliqu\'es \`a partir de s\'equences de Sobol'. Les propri\'et\'es de remplissement de l'espace des plans construits sont \'etudi\'ees sur la base de deux crit\`eres.
{\it Pour citer cet article~: L. Gilquin, Ll. A. Jim\'enez Rugama, E. Arnaud, F. J. Hickernell, H. Monod, C. Prieur, C. R. Acad. Sci.
Paris, Ser. I 340 (2005).}
\end{abstract}
\end{frontmatter}
\selectlanguage{english}
\section{Introduction}
Mathematical models often involve a substantial number of poorly known parameters. The effect of these parameters on the output of the model can be assessed through sensitivity analysis. Global sensitivity analysis methods are useful tools to identify the parameters having the most influence on the output. A well known approach is the variance based method introduced by Sobol' in \cite{Sobol2}. This method estimates sensitivity indices called Sobol' indices that summarize the influence of each model input. Among all Sobol' indices, one can distinguish first-order indices that estimate the main effect of each input.

@@

\subsection{Preliminaries}
Digital nets and sequences were first introduced by Niederreiter \cite{Niederreiter} in the numerical integration framework to define good uniformly distributed points in $\cube$. They can also appear in the literature as digital $(t,m,s)$-nets and digital $(t,s)$-sequences, or simply $(t,m,s)$-nets and $(t,s)$-sequences. Sobol' and Niederreiter-Xing sequences are two examples of digital sequences detailed in \cite{Sobol1} and \cite{NiederreiterXing}.

@@

Consider the generating matrices $C_1,\dots,C_s$, and $C_1^m,\dots,C_s^m$ their upper left corner blocks of size $m\times m$. Although $C_1,\dots,C_s$ are of infinite size, one only requires the knowledge of $C_1^m,\dots,C_s^m$ to construct the first $2^m$ points: for each $i=0,\dots,2^m-1$, the point $\vx_i = (x_{i,1},\dots,x_{i,s})^\intercal$ of the sequence is obtained dimension-wise by:
\begin{equation}
\label{dig.net.eq.}
(x_{i,j,1},\dots,x_{i,j,m})^\intercal = C_j^{m} \vi,\qquad j= 1,\dots,s\, ,
\end{equation}
where $x_{i,j} = \sum_{k \geq 1}^m x_{i,j,k}2^{-k}$ is the binary expansion of $x_{i,j}$ and $\vi = (i_{0},\dots,i_{m-1})^\intercal$ is the vector obtained from the binary expansion of $i= \sum_{k \geq 0}^{m-1} i_{k}2^{k}$. All matrix operations are performed in $\mathbb{F}_2$. Below we provide an example on how to compute $x_{11,1}=0.8125$ and $x_{9,2}=0.4375$,
\begin{equation*}
\begin{pmatrix}
x_{11,1,1} \\
x_{11,1,2} \\
x_{11,1,3} \\
x_{11,1,4}
\end{pmatrix} = 
\overbrace{
\begin{pmatrix}
1 & 0 & 0 & 0 \\
0 & 1 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1
\end{pmatrix}}^{C_1^4}
\overbrace{
\begin{pmatrix}
1 \\
1 \\
0 \\
1
\end{pmatrix}}^{\bvec{11}}=
\begin{pmatrix}
1 \\
1 \\
0 \\
1
\end{pmatrix}
,\qquad
\begin{pmatrix}
x_{9,2,1} \\
x_{9,2,2} \\
x_{9,2,3} \\
x_{9,2,4}
\end{pmatrix} = 
\overbrace{\begin{pmatrix}
1 & 1 & 1 & 1 \\
0 & 1 & 0 & 1 \\
0 & 0 & 1 & 1 \\
0 & 0 & 0 & 1
\end{pmatrix}}^{C_2^4}
\overbrace{\begin{pmatrix}
1 \\
0 \\
0 \\
1
\end{pmatrix}}^{\bvec{9}} =
\begin{pmatrix}
0 \\
1 \\
1 \\
1
\end{pmatrix}
\end{equation*}

@@

\begin{lemma}\label{Sobol_replicated}
All Sobol' sequences are digitally replicated of order 1.
\end{lemma}
\textit{Proof}: Consider any two $s$-dimensional Sobol' sequences generated by $C_1,\dots,C_s$ and $C'_1,\dots,C'_s$ respectively. Since any two matrices $C^{m}_j$ and ${C'}^{m}_j$ are square and full rank, the operations $\vi \mapsto C^{m}_j\vi$ and $\vi \mapsto {C'}^{m}_j\vi$ are one-to one and onto for all $i=0,\dots,2^m-1$. Therefore, they generate the same point sets.

@@

Two replicated point sets of order 1 $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ can be constructed using two $s$-dimensional Sobol' sequences. We note $C_1,\dots,C_s$ the generating matrices used to generate $\mathcal{P}_\ell$ and ${C'}_1,\dots,{C'}_s$ those used to generate $\mathcal{P}'_\ell$. To insure that $2$-dimensional projections between $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ are as uniform as possible, these generating matrices need to be different from each other. Our choice of the $2s$ generating matrices $C_1,\dots,C_s,{C'}_1,\dots,{C'}_s$ follows the results obtained by Joe and Kuo in \cite{Kuo}.
\bigskip

@@

With the multiplicative approach, the size of designs $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ is multiplied by $2$ each time $\ell$ is increased by one. This growth rate may be inadequate for some applications. The additive approach presented in this section is attractive due to a slower size growth. Given an initial choice of $r\geq 1$ that specifies the size of $B_0$ and ${B'}_0$, only $|B_0|=2^r$ points are added to both designs at each step. The main drawback of this approach is that $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ do not possess the structure of a Sobol' sequence anymore when $m\geq 1$. Still, both designs display good space-filling properties as shown in the next section.

@@

The two replicated point sets $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ constructed with the additive approach are iteratively refined with $B_\ell$ and ${B'}_\ell$. Each $B_\ell$ and ${B'}_\ell$ are $(t,r,s)$-nets. First, $B_0=\{\vx_i^{(0)}\}_{i=0}^{2^r-1}$ and ${B'}_0=\{{\vx'}_i^{(0)}\}_{i=0}^{2^r-1}$ are set to be the first $2^r$ Sobol' points of two $s$-dimensional sequences. The generating matrices of these two sequences are selected as in the multiplicative approach. Then, $B_\ell$ (resp. ${B'}_\ell$), for $\ell \geq 1$, is obtained from $B_0$ (resp. ${B'}_0$) by carrying out digital shifts and scrambling operations. These operations guarantee that both $B_\ell$ and ${B'}_\ell$ inherit the $(t,r,s)$-net structure of the initial sets $B_0$ and ${B'}_0$. We detail below the construction of $B_\ell$ and ${B'}_\ell$.
\bigskip

@@

The left multiplications carried out in Equations (\ref{dig.shift}) and (\ref{ibinom.scrambling}) by matrices $L$ and $L'$ are referred as random linear scrambling \cite{Matousek} while the additions by vectors $\ve_j$ and ${\ve'}_j$ are referred as digital shifts. There are exactly $2^r$ different choices for vectors $\ve_j$ and ${\ve'}_j$ and $2^{r(r-1)/2}$ choices for matrices $L$ and $L'$. The vectors $\ve_j$ (resp. ${\ve}'_j$) are selected as follows:
\begin{itemize}
\item[i)] for the first coordinate, the same vector $\ve_1$ (resp. ${\ve}'_1$) is selected at each iteration $\ell$,
\item[ii)] for the others coordinates, the vectors $\ve_2,\dots,\ve_s$ (resp. ${\ve}'_2,\dots,{\ve}'_s$) are selected to guarantee that the set $\{\ve_2,\dots,\ve_s\}$ is different at each step $\ell$. There exists exactly $2^{r (s-1)}$ combinations of this set which leaves us with a large panel of choice.
\end{itemize}

@@

\begin{lemma}
Using the additive construction, $\mathcal{P}_\ell$ and $\mathcal{P}'_\ell$ are two replicated designs of order $1$ for $\ell \geq 0$.
\end{lemma}
\textit{Proof}:For each $\ell\geq 1$, digital shift and scrambling operations carried out in equations (\ref{dig.shift}) and (\ref{ibinom.scrambling}) are bijections from $\mathbb{F}^r_2$ to $\mathbb{F}^r_2$. Hence, same coordinates of $B_\ell$ and ${B'}_\ell$ will contain the same set of $2^r$ values.

@@

We denote by design $A$ (resp. design $B$) one of the two replicated designs constructed with the multiplicative (resp. additive) approach. Values of maximin and discrepancy criteria are computed for designs $A$ and $B$. For design $A$, the design constructed is refined until reaching $2^{12}$ points. For design $B$, we chose a starting value $r=8$ and refine the design $15$ times. To highlight the performance of design $B$, the results are compared to those obtained with a Latin Hypercube design. Averaged results over $100$ repeats are presented in Figure \ref{results.crit}. The x-axis corresponds to the growing size of the designs from $2^8$ points to $2^{12}$ points.

As expected design $A$ performs the best. However, we observe that design $B$ outperforms a classic Latin Hypercube design for both criteria while allowing $15$ refinement steps from $2^8$ to $2^{12}$ points instead of $4$ steps for design $A$.

\begin{figure}
\caption{Averaged $L_2$ star discrepancy (left figure) and maximin (right figure) values for designs $A$ and $B$ over $100$ repeats.}
\label{results.crit}
\centering
\includegraphics[scale=0.4]{discrep.pdf}
\hspace*{0.1cm}
\includegraphics[scale=0.4]{mindist.pdf}
\end{figure}
